### 2013-07-23 doc/dissertation/document/chapters/mc_background/mc_background.tex
finished add step problem to nonlinear chapter
document formatting for appearance

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \citep{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \citep{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic-Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.

@@

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method with the collision estimator as
compared to the direct method. To assess both the CPU time and number
of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points in
the system. Solver parameters were set to a weight cutoff of
\sn{1}{-4} for the stochastic linear solver and a convergence
tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and the remaining sections of
this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the adjoint and direct
      solvers are used with the five-point and nine-point stencils.}}
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note
on the semi-log plot that we are indeed achieving the expected
exponential convergence from MCSA with both Monte Carlo
solvers. Second, we note that using the adjoint method with the same
number of stochastic histories per MCSA iteration gives a faster rate
of converge for the same reasons as above. We also note here that
fewer iterations are required for convergence when the 9-point stencil
is used to discretize the Laplacian operator (although at no gain in
speed as given by the results in
Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
smaller discretization error directly corresponds to a more well
defined residual source generated by the Richardson extrapolation for
the Monte Carlo calculation. In addition, the better defined source is
transported through a domain described more accurately by the 9-point
stencil, thus yielding a more accurate correction vector from the
Monte Carlo calculation.
\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/seq_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver.}}
  \label{fig:seq_iterations}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_100.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=100$.}
    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
      with the five point stencils and the adjoint Monte Carlo
      solver.}}
  \label{fig:seq_100}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/seq_conv_500.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
      with the five point stencils and the adjoint Monte Carlo
      solver.}}
  \label{fig:seq_500}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
      histories per MCSA iteration for a $200 \times 200$ square mesh
      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is
      significantly better than the collision estimator. At higher
      numbers of histories, the estimators become roughly
      equivalent.}}
  \label{fig:estimator_nh_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_nh_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
      per MCSA iteration for a $200 \times 200$ square mesh and a
      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is better
      than the collision estimator due to a lower iteration count
      while the actual compute time per iteration is higher.}}
  \label{fig:estimator_nh_time}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. history weight
      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is significantly better than the collision
      estimator. At higher numbers of histories, the estimators become
      roughly equivalent.}}
  \label{fig:estimator_wc_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/estimator_wc_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
      for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is better than the collision estimator due to a
      lower iteration count while the actual compute time per
      iteration is higher.}}
  \label{fig:estimator_wc_time}
\end{figure}

@@

\begin{figure}[p!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. artificial
      absorption probability for a $200 \times 200$ square mesh and
      40,000 histories.}  \textit{The addition of absorption is
      detrimental to iterative performance due to both the shortening
      of the random walks as well as the modification of the
      transition weight.}}
  \label{fig:absorption_iters}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=6in,clip]{chapters/mc_background/absorption_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
      probability for a $200 \times 200$ square mesh and 40,000
      histories.}  \textit{The reduced random walk length speeds up
      the calculation with a fixed number of histories. Too much
      artificial absorption increases iterations rapidly giving an
      increasing CPU time.}}
  \label{fig:absorption_time}
\end{figure}
\clearpage

@@

In this chapter, the fundamentals of Monte Carlo synthetic
acceleration methods were presented and explored in the context of a
simple model transport system. The following are the signficant
observations and findings.

\begin{itemize}
\item Monte Carlo synthetic acceleration (MCSA) has been generalized
  for all discrete linear transport problems
\item MCSA can be preconditioned using a general left/right scheme but
  requires the explicit inversion of the preconditioners and
  construction of the composite linear operator
\item The adjoint Neumann-Ulam method performs better when used with
  MCSA compared to the forward method
\item MCSA has superior performance to Halton's residual Monte Carlo
  method
\item The Richardson iteration in MCSA behaves as a smoother for the
  stochastic Monte Carlo error
\item MCSA iterative performance when used with the expected value
  estimator is insensitive to histories per iteration
\item MCSA iterative performance is insensitive to the stochastic
  history weight cutoff
\item Variance reduction with artificial absorption can enhance time
  to solution
\item The reduced domain approximation was introduced to mitigate the
  consequences of composite linear operator generation
\end{itemize}

>>>>

%%---------------------------------------------------------------------------%%
\section{Monte Carlo Synthetic Acceleration}
\label{sec:mcsa}
Using the ideas of Halton, Evans and Mosher recently developed a Monte
Carlo solution method that was not prohibited severely by the quality
of the initial guess for the system \citep{evans_monte_2009} and later
applied it more rigorously as a solution mechanism for the radiation
diffusion equation \citep{evans_monte_2012}. With their new methods,
they achieved identical numerical results as conventional Krylov
solvers as well as comparable performance in both number of iterations
and CPU time. Their approach was instead to use residual Monte Carlo
as a synthetic acceleration for a stationary method. To derive this
method, we begin by splitting the operator in
Eq~(\ref{eq:linear_problem})
\begin{equation}
  \ve{x} = (\ve{I} - \ve{A})\ve{x} + \ve{b}\:.
  \label{eq:linear_split}
\end{equation}
With this we can then define the stationary method
\textit{Richardson's iteration} as:
\begin{equation}
  \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,
  \label{eq:richardsons_iteration}
\end{equation}
which will converge if $\rho(\ve{I} - \ve{A}) < 1$. We then define the
solution error at the $k^{th}$ iterate relative to the true solution:
\begin{equation}
  \delta \ve{x}^k = \ve{x} - \ve{x}^k\:.
  \label{eq:mcsa_error}
\end{equation}
Subtracting Eq~(\ref{eq:richardsons_iteration}) from
Eq~(\ref{eq:linear_split}) we get:
\begin{equation}
  \delta \ve{x}^{k+1} = (\ve{I} - \ve{A})\delta \ve{x}^k\:.
  \label{eq:mcsa_setup_1}
\end{equation}
Subtracting from this $(\ve{I} - \ve{A})\delta \ve{x}^{k+1}$ yields:
\begin{equation}
  \begin{split}
    \ve{A}\delta \ve{x}^{k+1} &= (\ve{I} -
    \ve{A})(\ve{x}^{k+1}-\ve{x}^{k}) \\ &= \ve{r}^{k+1}\:.
    \label{eq:mcsa_setup_2}
  \end{split}
\end{equation}
Using this, we define the following scheme that will converge in one
iteration if $\ve{A}$ is inverted exactly:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1} = (\ve{I} - \ve{A})\ve{x}^k + \ve{b}\:,\\
    \ve{A} \delta \ve{x}^{k+1} = \ve{r}^{k+1}\:,\\
    \ve{x} = \ve{x}^{k+1} + \delta \ve{x}^{k+1}\:.
  \end{gather}
  \label{eq:mcsa_setup_3}
\end{subequations}
However, $\ve{A}$ is only approximately inverted by our numerical
methods and therefore we instead pose an iterative scheme in which the
Monte Carlo solvers are used to invert the operator. The
\textit{Fixed-Point Monte Carlo Synthetic Acceleration} (MCSA) method
is defined as:
\begin{subequations}
  \begin{gather}
    \ve{x}^{k+1/2} = \ve{x}^k + \ve{r}^k\:,\\
    \ve{r}^{k+1/2} = \ve{b} - \ve{A}\ve{x}^{k+1/2}\:,\\
    \ve{A}\delta\ve{x}^{k+1/2} = \ve{r}^{k+1/2}\:,\\
    \ve{x}^{k+1} = \ve{x}^{k+1/2} + \delta \ve{x}^{k+1/2}\:,\\
    \ve{r}^{k+1} = \ve{b} - \ve{A}\ve{x}^{k+1}\:,
  \end{gather}
  \label{eq:mcsa}
\end{subequations}
where a Neumann-Ulam Monte Carlo method is used to generate the
solution correction from the residual and Richardson's iteration in
the first step has been rewritten as a residual correction. Using
Monte Carlo in this way achieves the same effect as Halton's method,
decoupling its convergence rate from the overall convergence rate of
the method. Here, the approximate Monte Carlo solution is not driven
to a particular convergence as it merely supplies a correction for the
initial guess generated by Richardson's iteration. Rather, only a set
number of histories are required using the Neumann-Ulam method to
generate the correction. In addition, the fact that the scheme in
Eq~(\ref{eq:mcsa_setup_3}) will converge in one iteration if $\ve{A}$
is inverted exactly means that as more and more stochastic histories
are used to compute the correction and the error is reduced towards
zero, the number of iterations required for MCSA to converge should
decrease accordingly, thus accelerating the solution.

@@

A timing and convergence study is used to demonstrate the
effectiveness of the adjoint method with the collision estimator as
compared to the direct method. To assess both the CPU time and number
of iterations required to converge to a solution, a problem of
constant $\Delta$ was used with varying values of the number of mesh
elements, fixing the spectral radius of the system at a constant value
for each variation. Both the five-point and nine-point stencils were
used with both the direct and adjoint solvers. For each case, $N
\times N$ total random walk permutations were computed per MCSA
iteration where $N \times N$ is the number of discrete grid points in
the system. Solver parameters were set to a weight cutoff of
\sn{1}{-4} for the stochastic linear solver and a convergence
tolerance of \sn{1}{-8} for the MCSA iterative solver.
Figure~\ref{fig:poisson_cpu_time} gives the CPU time needed for each
case to converge in seconds and Figure~\ref{fig:poisson_iterations}
gives the number of iterations needed for each case to converge to the
specified tolerance as a function of the problem size. All
computations presented in this section and the remaining sections of
this chapter were completed on a 3.0 GHz Intel Core 2 Quad Q9650 CPU
machine with 16 GB 1067 MHz DDR3 memory.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the adjoint and
      direct solvers are used with the five point and nine point
      stencils. A CPU time speedup is noted with the adjoint method
      due to the higher density of random walk events in regions with
      a large residual.}}
  \label{fig:poisson_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the adjoint and direct
      solvers are used with the five-point and nine-point stencils.}}
  \label{fig:poisson_iterations}
\end{figure}

@@

As an additional comparison, the convergence behavior of MCSA can be
analyzed using both the adjoint and direct solvers to detect any
performance benefits. To assess the convergence properties of MCSA
using each solver and stencil, the infinity norm of the residual
computed in Eq.~(\ref{eq:mcsa}) was collected at each iteration for a
fixed problem size of $N=500$. Figure~\ref{fig:poisson_convergence}
gives the results of these computations. First, it is worthy to note
on the semi-log plot that we are indeed achieving the expected
exponential convergence from MCSA with both Monte Carlo
solvers. Second, we note that using the adjoint method with the same
number of stochastic histories per MCSA iteration gives a faster rate
of converge for the same reasons as above. We also note here that
fewer iterations are required for convergence when the 9-point stencil
is used to discretize the Laplacian operator (although at no gain in
speed as given by the results in
Figure~\ref{fig:poisson_cpu_time}). This is due to the fact that the
smaller discretization error directly corresponds to a more well
defined residual source generated by the Richardson extrapolation for
the Monte Carlo calculation. In addition, the better defined source is
transported through a domain described more accurately by the 9-point
stencil, thus yielding a more accurate correction vector from the
Monte Carlo calculation.
\begin{figure}[t!]
  \centering
  \includegraphics[width=5in,clip]{chapters/mc_background/dir_adj_conv.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the adjoint and direct solvers are used with the five
      point and nine point stencils. A higher rate of convergence is
      observed for MCSA using the adjoint Monte Carlo solver as
      compared to the direct method when both solvers compute the same
      number of random walks per iteration.}}
  \label{fig:poisson_convergence}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_cpu.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Problem Size ($N$ for
      an $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver. The number of random walks was
      twice the number of discrete states in the system in order to
      ensure convergence in the Sequential Monte Carlo method.}}
  \label{fig:seq_cpu_time}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_iterations.pdf}
  \caption{\textbf{Iterations to converge vs. Problem Size ($N$ for an
      $N \times N$ square mesh).} \textit{Both the Sequential Monte
      Carlo and MCSA solvers are used with the five point stencils and
      the adjoint Monte Carlo solver.}}
  \label{fig:seq_iterations}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_100.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=100$.}
    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
      with the five point stencils and the adjoint Monte Carlo
      solver.}}
  \label{fig:seq_100}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/seq_conv_500.pdf}
  \caption{\textbf{Infinity norm of the solution residual
      vs. iteration number for a problem of size $N=500$.}
    \textit{Both the Sequential Monte Carlo and MCSA solvers are used
      with the five point stencils and the adjoint Monte Carlo
      solver.}}
  \label{fig:seq_500}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. Monte Carlo
      histories per MCSA iteration for a $200 \times 200$ square mesh
      and a weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is
      significantly better than the collision estimator. At higher
      numbers of histories, the estimators become roughly
      equivalent.}}
  \label{fig:estimator_nh_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_nh_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. Monte Carlo histories
      per MCSA iteration for a $200 \times 200$ square mesh and a
      weight cutoff of \sn{1}{-4}.} \textit{For low numbers of
      histories, the expected value estimator performance is better
      than the collision estimator due to a lower iteration count
      while the actual compute time per iteration is higher.}}
  \label{fig:estimator_nh_time}
\end{figure}

@@

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. history weight
      cutoff for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is significantly better than the collision
      estimator. At higher numbers of histories, the estimators become
      roughly equivalent.}}
  \label{fig:estimator_wc_iters}
\end{figure}

\begin{figure}[t!]
  \centering
  \includegraphics[width=4.75in,clip]{chapters/mc_background/estimator_wc_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. history weight cutoff
      for a $200 \times 200$ square mesh and 40,000 histories.}
    \textit{For low numbers of histories, the expected value estimator
      performance is better than the collision estimator due to a
      lower iteration count while the actual compute time per
      iteration is higher.}}
  \label{fig:estimator_wc_time}
\end{figure}

@@

\begin{figure}[p!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_iters.pdf}
  \caption{\textbf{Iterations (s) to converge vs. artificial
      absorption probability for a $200 \times 200$ square mesh and
      40,000 histories.}  \textit{The addition of absorption is
      detrimental to iterative performance due to both the shortening
      of the random walks as well as the modification of the
      transition weight.}}
  \label{fig:absorption_iters}
\end{figure}

\begin{figure}[p!]
  \centering
  \includegraphics[width=4.5in,clip]{chapters/mc_background/absorption_time.pdf}
  \caption{\textbf{CPU Time (s) to converge vs. artificial absorption
      probability for a $200 \times 200$ square mesh and 40,000
      histories.}  \textit{The reduced random walk length speeds up
      the calculation with a fixed number of histories. Too much
      artificial absorption increases iterations rapidly giving an
      increasing CPU time.}}
  \label{fig:absorption_time}
\end{figure}
\clearpage

@@

In this chapter, the fundamentals of Monte Carlo Synthetic
Acceleration methods were presented and explored in the context of a
simple model transport system. The following are the significant
observations and findings.

\begin{itemize}
\item Monte Carlo Synthetic Acceleration (MCSA) has been generalized
  for all discrete linear transport problems
\item MCSA can be preconditioned using a general left/right scheme but
  requires the explicit inversion of the preconditioners and
  construction of the composite linear operator
\item The adjoint Neumann-Ulam method performs better when used with
  MCSA compared to the forward method
\item MCSA has superior performance to Halton's residual Monte Carlo
  method
\item The Richardson iteration in MCSA behaves as a smoother for the
  stochastic Monte Carlo error
\item MCSA iterative performance when used with the expected value
  estimator is insensitive to histories per iteration
\item MCSA iterative performance is insensitive to the stochastic
  history weight cutoff
\item Variance reduction with artificial absorption can enhance time
  to solution
\item The reduced domain approximation was introduced to mitigate the
  consequences of composite linear operator generation
\end{itemize}

### 2013-07-23 doc/dissertation/document/chapters/nonlinear_problem/nonlinear_problem.tex
document formatting for appearance

In this chapter we have developed and explored Monte Carlo synthetic
acceleration methods in the context of solutions to the Navier-Stokes
equations. The following are the significant observations and findings.

\begin{itemize}
\item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
  method leveraging Monte Carlo synthetic acceleration, has been
  developed
\item The FANM method has been incorporated into the Drekar
  multiphysics production code base developed at Sandia National
  Laboratories
\item The FANM method has been verified to produce the same solutions
  as a production Newton-Krylov method for three difficult benchmark
  problems for the Navier-Stokes equations in different flow regimes
  and geometries
\item The FANM method has better iterative performance than the
  Newton-Krylov method for convection dominated problems, converging
  in fewer linear solver iterations with the same preconditioning for
  high and low Rayleigh numbers
\item The Newton-Krylov method has better iterative performance than
  the FANM method for flow dominated by inertial forces, converging in
  fewer linear solver iterations with the same preconditioning for
  high Reynolds numbers
\item The spectral radius convergence restriction on MCSA was observed
  to be a significant hindrance by preventing solutions to forced flow
  problems at high Reynolds numbers
\item Explicit algebraic preconditioning used to overcome the spectral
  radius restriction of MCSA creates FANM run times
  $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
\item More Monte Carlo histories at every FANM iteration can reduce
  the number of linear and nonlinear iterations required to converge
  the problem
\item With future optimization and improved preconditioning, reactor
  accident analyses that rely on convection dominated transport or
  steady state calculations requiring driven flow could experience
  improved time to solution with FANM given the improved iterative
  performance observed for problems of these types
\end{itemize}

>>>>

In this chapter we have developed and explored Monte Carlo Synthetic
Acceleration methods in the context of solutions to the Navier-Stokes
equations. The following are the significant observations and findings.

\begin{itemize}
\item Forward-Automated Newton-MCSA (FANM), a new inexact Newton
  method leveraging Monte Carlo Synthetic Acceleration, has been
  developed
\item The FANM method has been incorporated into the Drekar
  multiphysics production code base developed at Sandia National
  Laboratories
\item The FANM method has been verified to produce the same solutions
  as a production Newton-Krylov method for three difficult benchmark
  problems for the Navier-Stokes equations in different flow regimes
  and geometries
\item The FANM method has better iterative performance than the
  Newton-Krylov method for convection dominated problems, converging
  in fewer linear solver iterations with the same preconditioning for
  high and low Rayleigh numbers
\item The Newton-Krylov method has better iterative performance than
  the FANM method for flow dominated by inertial forces, converging in
  fewer linear solver iterations with the same preconditioning for
  high Reynolds numbers
\item The spectral radius convergence restriction on MCSA was observed
  to be a significant hindrance by preventing solutions to forced flow
  problems at high Reynolds numbers
\item Explicit algebraic preconditioning used to overcome the spectral
  radius restriction of MCSA creates FANM run times
  $O(100)$-$O(1,000)$ slower than the Newton-Krylov solver
\item More Monte Carlo histories at every FANM iteration can reduce
  the number of linear and nonlinear iterations required to converge
  the problem
\item With future optimization and improved preconditioning, reactor
  accident analyses that rely on convection dominated transport or
  steady state calculations requiring driven flow could experience
  improved time to solution with FANM given the improved iterative
  performance observed for problems of these types
\end{itemize}
